---
title: "EDS212Resources"
author: "Adelaide Robinson"
date: '2022-08-05'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## R Markdown Setup

Add packages in the setup chunk\
the warning chunk sets warnings to not show up in the whole document

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
```

# R Markdown notes

adding a line break, two trailing spaces and then hit return

# Github

## Adding files to github

make sure you are in the terminal and not the console

-   gitadd -A (staging all the files)
-   git commit -m "my commit message"
-   git push
-   git status - check the status

## Adding to gitignore

click on the git ignore file

type nameofdoc.filetype for a specific file

\*.filetype for an entire file type

example: eds212.html, \*.html

## Working collaboratively in git

### Working without Branches

This workflow is effective if not working on the same files at the same
time. Also a good workfolow if you are working on different computers.

One person goes onto github and creates a project, and then add
collaborators, found under settings, collaborators

Collaborator should get a link, they can then clone the project, create
as a project on their computer and then make changes. when done push
their changes.Make sure to save before pushing.

The second person should be able to pull git pull and see those changes.

# GGplot

Add the tidyverse package in the setup chunk, and make sure data is
available

```{r}

#ggplot(data and mapping, anything to do with variables goes under aes) + type of plot
#basic scatterplot
ggplot(data = penguins, mapping = aes(x = body_mass_g, y = flipper_length_mm)) + geom_point(aes(color = species)) + labs(title = "Penguin", y = "Flipper length (mm)", x = "Body Mass (g)") 



```

# Functions

## Basic Function

```{r}
#first give the variables in the function, then define 
standard_weight <- function(a , b, L) {
  a * L^b }
  
#plug numbers into the function
great_barracuda <- standard_weight(a = 0.0181 , b = 3.27 , L = 60)
great_barracuda
```

## Logistic Growth Function

N~t~ = k/ 1 + (k - N~0~ / N~0~)e^-rt^ this function tells you the number
of individuals in a population given the initial population, carrying
capacity, the time and the rate

```{r}
#write this equation as a function r can use 
#reminder that to use e in r use exp()
pop_logistic <- function(carryingcap, init_pop, rate, time_yr) { carryingcap/ (1 + ((carryingcap - init_pop)/init_pop) * exp(- rate * time_yr))
  
}
#plug in given numbers to find exact population at a given time
pop_logistic(carryingcap = 2580, init_pop = 230, rate = 0.32, time_yr = 2.4)
# or create a sequence and show it over time
time_seq <- seq( from = 0, to = 20, by = 1)
#and then plug it in
chipmunk_pop <- pop_logistic(carryingcap = 2580, init_pop = 230, rate = 0.32, time_yr = time_seq)
#make this into a dataframe to graph
chip_dataframe <- data_frame(time_seq, chipmunk_pop)
#then graph
ggplot(data = chip_dataframe, aes(x = time_seq , y = chipmunk_pop)) + geom_point() + labs( title = "Chipmunk Population", x = "Time (yrs)", y = "Chipmunk Population Size")
```

## Solving the function for a sequence

```{r}
# make a vector with possible fish lengths, use the seq function and from, to and by
barracuda_length <- seq(from = 0 , to = 200 , by = 1)
#estimate weights using this vector
#plugging in known values for a and b
barracuda_weight <- standard_weight( a = 0.0181, b = 3.27 , L = barracuda_length)
#make these two vectors into a dataframe 
barracuda_dataframe <- data.frame(x = barracuda_length , y = barracuda_weight)
```
