---
title: "EDS212Resources"
author: "Adelaide Robinson"
date: '2022-08-05'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## R Markdown Setup

Add packages in the setup chunk the warning chunk sets warnings to not
show up in the whole document

```{r setup, include=FALSE}
#says nothing from this code chunk will show up 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
library(deSolve)
library(kableExtra)
library(here)
library(patchwork)
library(ggrepel)
library(gghighlight)
library(paletteer)
library(ggExtra)
library(ggbeeswarm)
library(gapminder)
library (sf)
```

# R Markdown notes

adding a line break, two trailing spaces and then hit return\
to add code in line, `r 1+1`

# Github

## setting up a project to be used on github

-   in the console type usethis::use_git make sure it is locally
    controlled

-   yes twice it will restart and be committed this is creating a git
    repository on our computer this still only exists on our computer

-   usethis::use_github, saves it on github, should open github

## Adding files to github

make sure you are in the terminal and not the console

-   gitadd -A (staging all the files)
-   git commit -m "my commit message"
-   git push
-   git status - check the status

## Adding to gitignore

click on the git ignore file

type nameofdoc.filetype for a specific file

\*.filetype for an entire file type

example: eds212.html, \*.html

note that if you don't do this from the very start you will need to
delete the file from github, this is the terminal command

    git rm file1.txt -F

## Working collaboratively in git

### Working without Branches

This workflow is effective if not working on the same files at the same
time. Also a good workflow if you are working on different computers.

One person goes onto github and creates a project, and then add
collaborators, found under settings, collaborators

Collaborator should get a link, they can then clone the project, create
as a project on their computer and then make changes. when done push
their changes.Make sure to save before pushing.

The second person should be able to pull git pull and see those changes.

## working with branches

creating a branch git checkout -B name of branch git push - u origin
new-branch (establishes remote repo) \# GGplot

Add the tidyverse package in the setup chunk, and make sure data is
available

# GG plot

## Scatter Plot

```{r}

#ggplot(data and mapping, anything to do with variables goes under aes) + type of plot
#Scatterplot
ggplot(data = penguins, mapping = aes(x = body_mass_g, y = flipper_length_mm)) + geom_point(aes(color = species)) + labs(title = "Penguin", y = "Flipper length (mm)", x = "Body Mass (g)") 


#reading in data 
lizards <- read_csv(here("lizards.csv"))
```

## box-plot

```{r}
#discrete variables on x, continuous on y
#middle line is the median
#lines of box at 1st and 3rd quartile
ggplot(data = penguins, aes(x = species, y = flipper_length_mm)) +geom_boxplot() + labs(title = "Penguin Flipper Length", y = "Species", x = "Flipper Length(mm)")
```

## Facet-Wrap

```{r}
ggplot(data=penguins, aes(x = flipper_length_mm, y = body_mass_g)) +geom_point(aes(color = species)) +facet_wrap(~island, ncol =1)
#add scales = free if you don't care about scales matching 

```

## Facet-Grid

\- \`facet_grid()\`: the one where you make a grid based on row & column
faceting variables

But what if we want to make a grid where the panels are split across
groups by lizard sex and if it has a broken tail or not? Since we have
two variables being used to create our grid, we'll use \`facet_grid()\`:

```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  facet_grid(sex ~ tail)
```

## Histogram

```{r}

#remember only one x varaible for a histogram, needs to be continous
ggplot(data = penguins, mapping = aes(x = bill_depth_mm)) + geom_histogram(color = "blue", aes(fill = species)) + labs(title = "Penguin Bill Depth", x = "Bill Depth (mm)", y = "Penguins") + theme_classic()
```

## Jitterplots

these are designed to separate values where a lot of data overlaps.
Doesn't look too different for this since the points do not extensively
overlap.

```{r}
ggplot(data = penguins, aes(x = flipper_length_mm, y= bill_depth_mm))+ geom_jitter()

```

## Beeswarm plot

```{r}
whiptails <- lizards |> 
  filter(common_name == "western whiptail") |> 
  drop_na(total_length, weight)

whiptail_bee <- ggplot(data = whiptails, aes(x = sex, y = weight)) +
  geom_beeswarm(size = 0.5) + geom_boxplot(fill = NA)


ggplot(data = whiptails, aes(x = sex, y = weight)) +
  geom_beeswarm(size = 0.5) + geom_boxplot(fill = NA)


```

## Marginal plot

```{r}
whiptail_plot <- whiptails |> 
  ggplot(aes( x = total_length, y = weight)) + geom_point(aes(color = sex)) +
  theme(legend.position = "bottom")


ggMarginal(whiptail_plot, type = "boxplot", groupColour = TRUE)
```

## Column Plot

```{r}
test<- penguins |> group_by(island, sex) |> summarize(total_penguins = n())

test |> ggplot(aes(x = island, y = total_penguins)) + geom_col()

#option to make this a proportion instead
test |> ggplot(aes(x = island, y = total_penguins, fill = sex)) + geom_col(position = "fill") + labs( y = "proportion of penguins", x = "island") +theme_minimal()
```

## Patchwork plot

```{r}
(whiptail_bee + whiptail_plot / whiptail_bee) #side by side

(whiptail_bee / whiptail_plot) #columns

```

## Heat map

```{r}
# Get the counts: 
lizard_counts <- lizards %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  count(year = lubridate::year(date), common_name) %>% 
  drop_na()

# Make a heatmap of counts:
ggplot(data = lizard_counts, aes(x = year, y = common_name)) +
  geom_tile(aes(fill = n), show.legend = FALSE) +
  geom_text(aes(label = n), color = "white", size = 3) +
  scale_fill_gradientn(colors = c("navy","red","orange")) +
  theme_minimal() +
  labs(x = "Year", y = "Lizard common name")
```

## Spatial Maps

use geom_sf

see eds 221 day 8 data_viz for example

## Factor Reorder

```{r}
lizard_counts <- lizards |> 
  count(common_name)

ggplot(data = lizard_counts, 
       aes(y = fct_reorder(common_name, n),# reorder common names by number
           x = n)) +
  geom_col()
```

## Titling graph

y axis variable vs x axis variable

```{r}
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g))+ geom_point(aes(color = species)) +labs(title = "Penguin body Mass vs Flipper Length", x = "Body Mass (g)", y = "Flipper Length (mm)", subtitle = "Palmer Penguins", caption = "EDS 212") + # manually choose colors
  scale_color_manual(values = c("pink", "red", "violet"))+
  #center labels
  theme(plot.title=element_text(hjust=0.5),
        plot.subtitle=element_text(hjust=0.5), plot.caption = element_text(hjust=0.5))
```

## Customization

Basic non aes changes

```{r}
lizards |> ggplot(aes(x = total_length)) + 
  geom_histogram(color = "orange",
                 size = 0.3,
                 fill = "purple",
                 linetype = "dotted"
                 )
#changes aesthetic not based on variable, don't put w/in aes


```

### Scaling Colors

We've learned to use \`scales\` functions to update breaks and labels.
It is also useful for updating aesthetics, for example to customize
color gradients, set size bins, and more.

Again, it is very important to know \*\*what aesthetic you are trying to
update.\*\*

For example, is it a \*\*fill\*\* aesthetic you're hoping to customize?
Then you might use \`scale_fill\_\*()\`. But if it is a \*\*color\*\*
aesthetic, you'd use \`scale_color\_\*()\` functions. If it is the
transparency, then \`scale_alpha\_\*()\`. If it is the shape, then
\`scale_shape\_\*()\`. So make sure you know \*which aesthetic you are
hoping to change with scales!\*

```{r}
#gradient
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_gradient(low = "purple", high ="orange")

#gradint with custom colors

ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_gradientn(
    colors = c("magenta", "cyan 4", "darkorchid3", "dodgerblue","green", "yellow"))

#step scale
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_steps(low = "red", high = "black")

#steps with middle color
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_steps2(
    low = "purple",
    mid = "white",
    high = "black",
    midpoint = 50,
    breaks = c(15, 30, 45))



ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = total_length)) +
  scale_color_stepsn(colors = c("orange",
                                "red",
                                "purple"),
                     breaks = seq(from = 0, to =300, by = 60))


```

### Color fills with palateer

```{r}
lizards_median <- lizards |> 
  mutate(common_name = fct_reorder(common_name, total_length,
                                   .fun = median))
#create a new column, overwrites existing column, recast common name as a factor 


ggplot(data = lizards_median, aes(y = common_name, x = total_length)) +
  geom_boxplot(aes(fill = common_name), show.legend = FALSE) +
  scale_fill_paletteer_d(palette = "nationalparkcolors::CraterLake")+
  theme_minimal()
 #view palateer options 
#view(paletteer::palettes_d_names)
```

### Options within Theme

```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) + 
  theme(panel.grid.major.x = element_line(color = "red"),
        panel.grid.minor.y = element_blank(), # leave it blank
        axis.title.x = element_text(color = "purple"),
        axis.text.y = element_text(color = "green"),
        text = element_text(size = ),
        panel.background = element_rect(color = "purple", fil = "yellow")) + annotate("text", x = 300, y = 50, label = "woooo", color = "green") +
  geom_vline(xintercept = 250, linetype = "dashed",
             color = "orange")
      
```

### ggrepel

```{r}
wws_lizards <- lizards |> filter(common_name == "western whiptail",
                      site == "sand")

#bad labels
ggplot(data = wws_lizards, aes(x = total_length, y = weight)) +
  geom_point()+
  geom_text(aes(label = toe_num))

#dodge points
ggplot(data = wws_lizards, aes(x = total_length, y = weight)) +
  geom_point()+
  geom_text_repel(aes(label = toe_num))


gapminder |> 
  filter(year == 2002, continent == "Europe") |> 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  geom_text_repel(aes(label = country, size = 2))

```

### gghighlight

```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  gghighlight(toe_num == 250, label_key = toe_num)+ #gets overwritten by next line
  gghighlight(weight >30, label_key = toe_num)
```

### Breaks and Limits

The important thing: know \*what type of variable you have on each
axis\* so that you know what \`scale\_\` version to call. For example:

\- For dates: \`scale\_\*\_date()\`

\- For continuous variables: \`scale\_\*\_continuous()\`

\- For discrete variables: \`scale\_\*\_discrete()\`

Within those layers added to your plot, you can update the \`breaks =
\`, \`limits = \`, \`labels = \` and \`expand = \`options.

```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point()

ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  scale_x_continuous(breaks = c(0, 250, 500), 
                     limits = c(0, 500)) +
  scale_y_continuous(breaks = seq(from = 0, to = 70, by = 10), 
                     limits = c(0, 70)) +
  theme_light()
```

convert to a log scale

```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  scale_x_log10()
```

formatting axis dates

```{r}
lizard_counts <- lizards %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  count(date)

ggplot(data = lizard_counts, aes(x = date, y = n)) +
  geom_line() +
  scale_x_date(breaks = scales::breaks_width("3 years"), # See date_breaks for next year!
               labels = scales::label_date("'%y")) + # See date_labels for next year!
  scale_y_log10(labels = scales::label_scientific())

ggplot(data = lizard_counts, aes(x = date, y = n)) +
  geom_line() +
  scale_x_date(breaks = scales::breaks_pretty())
```

### gridlines and panel colors

```{r}
p <- ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point()

p + 
  theme(panel.grid = element_blank()) # removes all gridlines (major & minor)

p + 
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "red"))

# Now let's just go bananas with some theme stuff: 

p + 
  theme(panel.background = element_rect(color = "purple", size = 3, fill = "yellow"),
        panel.grid.major.y = element_line(color = "orange"),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(color = "blue"),
        axis.text.y = element_text(color = "cyan"),
        axis.title.x = element_text(color = "green"),
        axis.title.y = element_text(color = "gray70"),
        text = element_text(size = 12, family="serif"),
        plot.background = element_rect(fill = "pink"))
```

### Direct annotation

```{r}
p +
  annotate("text", x = 100, y = 50, label = "COOL!", color = "purple") +
  annotate("text", x = 400, y = 25, label = "WHOA.", color = "red") +
  geom_hline(yintercept = 40, linetype = "dotted", color = "blue") +
  geom_vline(xintercept = 20, linetype = "dashed", color = "green") +
  theme_minimal()

# Or, have the value be determined based on a variable: 
p +
  geom_hline(yintercept = mean(lizards$weight), linetype = "dashed", color = "red") +
  annotate("text", x = 350, y = mean(lizards$weight) + 2, label = "Mean weight", color = "red")
```

# Functions

## Basic Function

```{r}
#first give the variables in the function, then define 
standard_weight <- function(a , b, L) {
  a * L^b }
  
#plug numbers into the function
great_barracuda <- standard_weight(a = 0.0181 , b = 3.27 , L = 60)
great_barracuda
```

## Logistic Growth Function

N~t~ = k/ 1 + (k - N~0~ / N~0~)e^-rt^ this function tells you the number
of individuals in a population given the initial population, carrying
capacity, the time and the rate

```{r}
#write this equation as a function r can use 
#reminder that to use e in r use exp()
pop_logistic <- function(carryingcap, init_pop, rate, time_yr) { carryingcap/ (1 + ((carryingcap - init_pop)/init_pop) * exp(- rate * time_yr))
  
}
#plug in given numbers to find exact population at a given time
pop_logistic(carryingcap = 2580, init_pop = 230, rate = 0.32, time_yr = 2.4)
# or create a sequence and show it over time
time_seq <- seq( from = 0, to = 20, by = 1)
#and then plug it in
chipmunk_pop <- pop_logistic(carryingcap = 2580, init_pop = 230, rate = 0.32, time_yr = time_seq)
#make this into a dataframe to graph
chip_dataframe <- data_frame(time_seq, chipmunk_pop)
#then graph
ggplot(data = chip_dataframe, aes(x = time_seq , y = chipmunk_pop)) + geom_point() + labs( title = "Chipmunk Population", x = "Time (yrs)", y = "Chipmunk Population Size")
```

## Solving the function for a sequence

```{r}
# make a vector with possible fish lengths, use the seq function and from, to and by
barracuda_length <- seq(from = 0 , to = 200 , by = 1)
#estimate weights using this vector
#plugging in known values for a and b
barracuda_weight <- standard_weight( a = 0.0181, b = 3.27 , L = barracuda_length)
#make these two vectors into a dataframe 
barracuda_dataframe <- data.frame(x = barracuda_length , y = barracuda_weight)
```

# Derivatives

Taking the derivative for an expression

```{r}
#define the expression
func1 <- expression(5 * x ^ 2)
#take the derivative, define the variable which you are taking the derivative in respect to
my_deriv <- deriv(func1, "x")
my_deriv
#define x to solve for a specific number
#second number is the answer
x <- 1
eval(my_deriv)

```

another option, gives simpler output

Find the slope of \$g(z) = 2 z\^3 -10.5z\^2 +4.1\$

```{r}
func2 <- expression(2 * z ^ 3 -10.5 *z ^2 +4.1)
your_deriv <- D(func2, "z")
your_deriv
#find derivative over a sequence of values
z <- seq(from = 0, to = 5, by = 0.5)
eval(your_deriv)
```

## ODE function & Lotka-Volterra Example

\*\*Credit:\*\* This code is closely based on the article [Numerically
solving differential equations with
R](%5Bhttps://rstudio-pubs-static.s3.amazonaws.com/32888_197d1a1896534397b67fb04e0d4899ae.html)](<https://rstudio-pubs-static.s3.amazonaws.com/32888_197d1a1896534397b67fb04e0d4899ae.html>))

As described in the lecture, the Lotke-Volterra models have been used to
describe predator-prey populations.

#### Prey equation:

$$\frac{dx}{dt}=\alpha x-\beta xy$$

From Wikipedia: "The prey are assumed to have an unlimited food supply
and to reproduce exponentially, unless subject to predation; this
exponential growth is represented in the equation above by the term
$alpha * x$. The rate of predation upon the prey is assumed to be
proportional to the rate at which the predators and the prey meet, this
is represented above by

$\beta xy$. If either x or y is zero, then there can be no predation."

$$\frac{dy}{dt}=\delta xy - \gamma y$$

From Wikipedia: "In this equation, $\delta xy$ represents the growth of
the predator population. (Note the similarity to the predation rate;
however, a different constant is used, as the rate at which the predator
population grows is not necessarily equal to the rate at which it
consumes the prey). The term $\gamma y$ represents the loss rate of the
predators due to either natural death or emigration, it leads to an
exponential decay in the absence of prey.

Where:

\- $x$ is prey population (e.g. rabbits)

\- $y$ is predator population (e.g. wolves)

\- $\alpha, \beta, \gamma, \delta$ are positive parameters

To find an approximate solution in R, we will need four things:

-   Parameter values

-   A sequence of times over which we'll approximate predator & prey
    populations -

-   An initial condition (initial populations of predator & prey at t =

    0)  

-   The differential equations that need to be solved

Solving the Lotke-Volterra equation:

```{r}
# Create a sequence of times (days): 
time <- seq(0, 25, by = 0.05)

# Set some parameter values (these can change - keep it in mind):
parameters <- c(alpha = .85, beta = 0.8, delta = 0.5, gamma = 1)

# Set the initial condition (prey and predator populations at t = 0).
# Recall: x = prey, y = predator
init_cond <- c(x = 5, y = 3)

# Prepare the series of differential equations as a function: 
lk_equations <- function(time, init_cond, parameters) {
  with(as.list(c(init_cond, parameters)), {
    dxdt = alpha * x - beta * x * y
    dydt = delta * x * y - gamma * y
    return(list(c(dxdt, dydt)))
  })
}

# Find the approximate the solution using `deSolve::ode()`:
approx_lk <- ode(y = init_cond, times = time, func = lk_equations, parms = parameters)

# Check the class: 
class(approx_lk)

# We really want this to be a data frame, and we want both prey (x) and predator (y) to be in the same column -- we'll learn why in EDS 221 (tidy data)
approx_lk_df <- data.frame(approx_lk) %>% 
  pivot_longer(cols = c(x,y), names_to = "species", values_to = "population")

# Plot it! 
ggplot(data = approx_lk_df, aes(x = time, y = population)) +
  geom_line(aes(color = species))



```

# Vectors, Matrices and dot product

## making vectors

```{r}
marmots <- c("blue", "green", 4, "yellow")

#Check the class

class(marmots)

pika <- c(12.4, 6.8, 2.9, 8.8, 8.5)

class(pika)
#scalar multiplier

scalar_pika <- 5.2 * pika

scalar_pika
#you can also add or subtract vectors ( as long as they are the same length)
ferret <- c(1, 12, 18, 2, 8)

pika - ferret

```

## dot product

The dot product is the sum of elements of each vector multiplied
together, and is a measure of how close the vectors "point" in the same
direction

```{r}
pika %*% ferret
```

\`\`\`

## Matrices

```{r}
my_values <- seq(from = 1, to = 10, length =200)
# this will fill the data into a matrix in order going by row
my_matrix <- matrix(data = my_values, nrow = 2, ncol = 5, byrow = TRUE)

5 * my_matrix

class(my_matrix)
```

Further info on matrices in eds 221

# Logical Expressions

```{r}
#if 5 =1 
5 == 1
"cat" == "dog"
# true if cat does not equal dog
"cat" != "dog"
# runs through the vector and checks for each
meds <- c(1,2,3,4,5)
meds == 3
meds == 1 | meds >3

teddy <- 5
banjo <- 7 
tallie <- 4


teddy <= banjo
```
